---
title: "Assignment_02"
output:
  pdf_document: default
  # html_notebook: default
---
1. Consider the built-in data frame `chickwts`.
    (a) How many variables in this data frame? What are the meanings of these variables? (Check the
    help file/ r documentation if needed.)
    (b) Draw a boxplot to illustrate the weight of different feed types. (Put all boxplots in a single plot.)
    (c) Create a subset of the data frame called chickwtsLinseed which contains all observations for
    which the chicks were fed linseed.
    (d) Calculate the average weight of the chicks who were fed linseed.
    (e) (Learn t.test() function by ?t.test if you have never used it.) Compare the weights of chicks
    in group linseed and group sunflower by a two-sample t test. What is the p value?
    (f) Find number of observations, mean, median and standard deviation of weight for each feed type.

```{r}
# How many variables in this data frame? What are the meanings of these variables?
q1_df <- chickwts
cat("Variable names are - '", paste(names(q1_df), collapse = "' and '"), "'", sep="")
head(q1_df,5)
```

```{r}
# Draw a boxplot to illustrate the weight of different feed types.
# (Put all boxplots in a single plot.)
par(cex.axis = 0.8) # Reduce the fontsize of the x and y ticks to make it fit
boxplot(q1_df$weight ~ q1_df$feed, col = "#c79cac",
        varwidth = FALSE, notch = FALSE, main = "Chicken Weight data",
        ylab = "Weight at six weeks (gm)", xlab = "Feed Type", boxwex = 0.6)
```

```{r}
# Create a subset of the data frame called chickwtsLinseed which contains all
# observations for which the chicks were fed linseed.
chickwtsLinseed <- q1_df[q1_df$feed == "linseed", ]
rownames(chickwtsLinseed) <- seq_len(nrow(chickwtsLinseed))    # Reset column index

# Calculate the average weight of the chicks who were fed linseed.
cat("Average weight for linseed chicks is", mean(chickwtsLinseed$weight), "gm.")
```

```{r}
# Compare the weights of chicks in group linseed and group sunflower by a
# two-sample t test. What is the p value?
chickwtsSunflower <- q1_df[q1_df$feed == "sunflower", ]
rownames(chickwtsSunflower) <- seq_len(nrow(chickwtsSunflower))
chickwtsTtest <- t.test(chickwtsLinseed$weight, chickwtsSunflower$weight)
chickwtsPval <- chickwtsTtest$p.value
cat("The p-value is", format(chickwtsPval, scientific = FALSE))
```

```{r}
# Find number of observations, mean, median and standard deviation of weight for each feed type
summary_stats_list <- tapply(q1_df$weight, q1_df$feed, function(x) {
  len_df <- length(x)
  mn_df <- mean(x)
  medn_df <- median(x)
  sd_df <- sd(x)
  agg <- c(num = len_df, mean = mn_df, median = medn_df, sd = sd_df)
  return(agg)
})

summary_df <- as.data.frame(do.call(rbind, summary_stats_list))
print(summary_df)
```

<br><br>

2. The dataset rnf6080.dat records hourly rainfall at a certain location in Canada, 1960 to 1980.
    (a) Assign the data set in rnf6080.dat to a data frame called rain.df. (There are some missing
    values marked by -999. Can you change them to NA?)
    (b) What are the names of the columns of rain.df?
    (c) Use the following command to re-label of columns of this data frame:
    names(rain.df) <- c("year", "month", "day", seq(0,23))
    (d) Create a new column call daily which is the sum of the 24 hourly columns.
    (e) Find average amounts of rain fall for each month.

```{r}
# Assign rnf6080.dat to dataframe and change -999 to NA
rain_df <- read.table('rnf6080.dat')
rain_df[rain_df == -999] <- NA
head(rain_df, 5)
```

```{r}
# What are the names of the columns of rain.df?
names(rain_df)
```

```{r}
# Re-label the columns of the data frame
names(rain_df) <- c("year", "month", "day", seq(0, 23))
names(rain_df)
```

```{r}
# Create daily daily column equal to sum of the 24 hourly columns
library(dplyr)
rain_df <- mutate(rain_df, daily = rowSums(rain_df[, 4:27], na.rm = TRUE))
# rain_df$daily <- rowSums(rain_df[, 4:27], na.rm = TRUE)
head(rain_df, 5)
```

```{r}
# Find average amounts of rain fall for each month.
grouped_month <- rain_df %>%
  group_by(month) %>% # year,
  summarize(monthly_avg = mean(daily, na.rm = TRUE), .groups = "drop") %>%
  arrange(month)

print(grouped_month)
```

3. us contagious diseases is a data set in R package dslabs. It contains 6 variables with 16065
observations. Mumps is one of 7 diseases in the column disease.
    (a) The data set us contagious diseases is in a data frame format. Covert it to a tibble format.
    (b) Use filter to find rows with disease Mumps and year between 1991 and 2000. Show the first 10
    rows of your result.
    (c) Find the top 5 states with the most Mumps cases over the 10 years from 1991 to 2000.
    (d) Based on the data in part (b), add a column called case per mil which calculates Mumps cases
    per population (in unit ‘cases per million people’).
    (e) Find the top 5 states with the most Mumps cases per population on average over the 10 years from
    1991 to 2000.

```{r}
library("dslabs")
# Covert the dataset to a tibble format
us_cd <- us_contagious_diseases
cat(dim(us_cd), "\n", names(us_cd))
us_cd <- as_tibble(us_cd)
head(us_cd, 5)
```

```{r}
# Use filter to find rows with disease Mumps and year between 1991
# and 2000. Show the first 10 rows of your result
us_cd_filt <- filter(us_cd, disease == "Mumps", (year >= 1991 & year <= 2000))
head(us_cd_filt, 10)
```

```{r}
# Find the top 5 states with the most Mumps cases over the 10 years from 1991 to 2000.
tm_top_5 <- us_cd_filt %>%
  group_by(state) %>%
  summarize(sum_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(sum_count))

head(tm_top_5, 5)
```

```{r}
# add a column called case per mil which calculates Mumps cases 
# per population (in unit ‘cases per million people’)
us_cd_filt_pop <- mutate(us_cd_filt, case_per_mil = (count / population))
head(us_cd_filt_pop, 2)
```

```{r}
# Find the top 5 states with the most Mumps cases per population on average over 
# the 10 years from 1991 to 2000.
cpm_top_5 <- us_cd_filt_pop %>%
  group_by(state) %>%
  summarize(mean_cpm = mean(case_per_mil, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_cpm))

head(cpm_top_5, 5)
```

4. Consider the built-in data frame pressure. There are two columns temperature and pressure.
    (a) Construct a scatterplot with pressure on the vertical axis and temperature on the horizontal
    axis. Are the variables related linearly or nonlinearly?
    (b) The graph of the following function passes through the plotted points reasonably well
    y = (0.168 + 0.007x)^20/3.
    The differences between the pressure values predicted by the curve and the observed pressure
    values are called residuals. Here is a way to calculate them:
    residuals <- with(pressure, pressure - (0.168 + 0.007*temperature)^(20/3))
    Construct a normal QQ plot of these residuals via qqnorm() and decide whether they are normally
    distributed or whether they follow a skewed distribution.
    (c) Now, apply the power transform y3/20 to the pressure data values. Plot these transformed values
    against temperature. Is a linear no nonlinear relationship evident now?
    (d) Calculate residuals for the difference between transformed pressure values and those predicted by
    the straight line. Obtain a normal QQ plot, and decide whether the residuals follow a normal
    distribution or not.

```{r}
# Construct a scatterplot of pressure and temperature
pres <- pressure
plot(pres$temperature, pres$pressure, xlab = "temperature", ylab = "pressure")
print("The variables are non-linearly correlated")
```

```{r}
# Fit a linear model using y = (0.168 + 0.007x)^20/3
# Estimate the residuals and construct a normal QQ plot
pres <- pres %>%
  mutate(model = (0.168 + (0.007 * temperature))^(20 / 3)) %>%
  mutate(residual = pressure - model)
# plot(pres$temperature, pres$pressure, xlab = "temperature", ylab = "pressure")
# lines(pres$temperature, pres$model, type = "l", col = "red")
qqnorm(pres$residual)
print("The data follows a skewed distribution")
```

```{r}
# Apply the power transform y^(3/20) to the pressure data values and
# plot it against temperatures
pres <- pres %>%
  mutate(pres_trans = pressure^(3 / 20)) %>%
  mutate(residual_t = pres_trans - (0.168 + 0.007 * temperature))

plot(pres$temperature, pres$pres_trans, xlab = "temperature",
     ylab = "transformed pressure")
print("A linear relationship is evident")
```

```{r}
qqnorm(pres$residual_t)
print("The residuals follow a normal distribution")
```

5. diamonds is a dataset in ggplot2 package. Read the R document on this dataset.
    (a) Create a new categorical variable ‘size’ based on the ‘carat’ (Small:<0.5, Medium: 0.5 to 1, Large:>1).
    (b) Group the data by ‘cut’ and ‘color’, and calculate the average price and carat for each group.
    (c) Create a facetted scatter plot of ‘price’ versus ‘carat’, facetted by ‘cut’, and colored by ‘clarity’.
    (d) Build a histogram of ‘price’, with each bar colored according to the ‘color’ category in the dataset.

```{r}
library(ggplot2)
dia <- diamonds
head(dia, 5)
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```