---
title: "Assignment_09_JO"
subtitle: "Markov Chain Monte Carlo Methods."
output:
  pdf_document: default
header-includes:
  - \usepackage{amsmath}
---

```{r}
set.seed(675)
```
Please answer the following questions and submit both R codes and the results. Make all results repro- ducible.

1. Generate $N(0,1)$ using random walk Metropolis algorithm based on proposal distribution 
    $g(x|x^{(t)}) \sim U(x^{(t)} - a, x^{(t)} + a)$. Generate Markov chains with length 20,000 using 
    $a = 0.1,\ 1\ and\ 10$, respectively. Compare the means, variances and distributions.
```{r}
```

2. Use independent sampler method (Independent Metropolis-Hastings Algorithm) to solve Example 9.5 in the 
  textbook (Statistical Computing with R page 256, available from UA library web). Use a flat 
  prior $U(0,0.5)$ for $\beta$. Use the same data as in the text book, i.e., (82, 72, 45, 34, 17), 
  so the results are comparable.
```{r}
```

3. Textbook Exercise 9.6.
```{r}
```

4.  Textbook Exercise 9.8. (Compare the empirical marginal distributions with the theoretical ones.).
```{r}
```