---
title: "Assignment_10_JO"
subtitle: "Optimization and EM Algorithm."
output:
  pdf_document: default
header-includes:
  - \usepackage{amsmath}
---

```{r}
set.seed(675)
library(ggplot2)
library(gridExtra)
```
Please answer the following questions and submit both R codes and the results. Make all results repro- ducible.

1. Implement secant method and use it to find the maximiser of $f(x) = \frac {log x} {1+x}$ by 
finding the root of $f'(x)$.
```{r}
# Define the function f(x)
f <- function(x) {
  return(log(x) / (1 + x))
}

# Define the derivative of f(x)
f_prime <- function(x) {
  return((1 - log(x) + x) / (1 + x)^2)
}

# Secant method to find the root of a function
secant_method <- function(f, x0, x1, tol = 1e-8, max_iter = 1000) {
  for (i in 1:max_iter) {
    x_next <- x1 - f(x1) * (x1 - x0) / (f(x1) - f(x0))
    if (abs(x_next - x1) < tol) {
      return(x_next)
    }
    x0 <- x1
    x1 <- x_next
  }
  warning("Maximum number of iterations reached")
  return(x_next)
}

# Find the root of f'(x)
root <- secant_method(f_prime, 1.0, 2.0)

# Maximiser of f(x) is where f'(x) = 0, so the value of x is the root we found
maximizer <- root

# Check if maximizer is valid
if (maximizer > 0) {
  cat("Maximizer of f(x) is:", maximizer, "\n")
  cat("Maximum value of f(x) is:", f(maximizer), "\n")
} else {
  cat("Maximizer not found in the domain of f(x)\n")
}

```

2. Define $f(x) = 2 log x + \sqrt {log(x + 1)} - x$.
    (a) Draw a plot of the function $f(x)$ over the interval (0, 10).
    (b) Use one of the build-in functions in R to find the maximizer of $f(x)$.
    (c) Implement by yourself one of the methods in Chapter 11 (bisection, Newton's, or secant 
    method) to find the maximizer of $f(x)$. Make sure that your maximizer is accurate up to 
    3 digits after the decimal point.
```{r}
# Define the function f(x)
f <- function(x) {
  return(2 * log(x) + sqrt(log(x + 1)) - x)
}

# Define the derivative of f(x)
f_prime <- function(x) {
  return((2 / x) + (sqrt(log(x + 1)) / (2 * (x + 1))) - 1)
}

# (a) Plot the function f(x) over the interval (0, 10)
x_values <- seq(0.01, 10, by = 0.01)  # Generate x values
y_values <- f(x_values)  # Calculate corresponding y values

plot(x_values, y_values, type = "l",
     col = "blue", xlab = "x", ylab = "f(x)",
     main = "Plot of f(x)")

# (b) Use uniroot function to find the maximizer of f(x)
result <- uniroot(f_prime, lower = 1, upper = 4)$root

# Round maximizer to 3 decimal places
max_value <- round(result, digits = 3)

# Print maximizer
cat("Maximizer of f(x) (up to 3 decimal places):", max_value, "\n")

# (c) Define the function for Newton's method

```

3. Consider the multinomial distribution with four outcomes
    $$ M(n, \frac {1}{2} + \frac {\theta}{4}, \frac {1}{4}(1-\theta), \frac {1}{4}(1-\theta), 
    \frac {\theta}{4}),\ \ \theta \in (0,1) $$
    (a) Given an observation $(x_1, x_2, x_3, x_4)$, write down the log-likelihood function $l(\theta)$.
    (b) If n = 197 and x = (125, 18, 20, 34), evaluate the MLE of $\theta$ by solving $l(\theta) = 0$. (You can solve it by hand.)
    (c) Write a program to find the maximizer of $l(\theta)$ by Newton's method, starting with $\theta^0 = 0.5$.
    (d) Write a program to find the maximizer of $l(\theta)$ by Fisher scoring approach (Lecture notes Page 210), starting with $\theta^0 = 0.5$.
    (e) Write a program to find the maximizer of $l(\theta)$ by the EM algorithm, starting with $\theta^0 = 0.5$.
    (Hint: If the first cell $x_1$ can be split into two cells $z_1, z_2$, we get multinomial distribution with
    five outcomes
    $$ (z_1, z_2, x_2, x_3, x_4)  M(n, \frac {1}{2} + \frac {\theta}{4}, \frac {1}{4}(1-\theta), \frac {1}{4}(1-\theta), 
    \frac {\theta}{4}),\ \ \theta \in (0,1) $$
    The log likelihood function is much easier, and the MLE can be calculated if $z_1$ and $z_2$ are known.)
    (f) Given a flat prior, approximate the posterior mean of $\theta$ using one of MCMC methods. (This is  essentially a 
    question from last assignment.)
    
    Note: Use n = 197 and x = (125, 18, 20, 34) when solving c-e. For c-e, write down the codes and 
    the iteration times needed to converge to the solution up to $10^{-4}$.
```{r}
```

4. Consider a sample of size n from a mixture distribution with unknown weights
    $$X_i ~ \theta N(0, 1) + (1 - \theta) N(2, 2),\ i = 1, 2, ..., n.$$
    (a) Describe an EM procedure to evaluate the MLE $\hat {\theta}$.
    (b) Calculate the MLE using the EM algorithm on a simulated dataset with $n = 25, \theta = 0.3$.
```{r}
# Print log and theta values
# Don't estimate the mean or variance parameters
```