---
title: "Assignment_03_JO"
output:
  pdf_document: default
  # html_notebook: default
---

```{r}
library(dplyr)
library(ggplot2)
library(microbenchmark)
library(Rcpp)
```

1. Construct a sequence as follows. Start with any positive integer n. Then each term is obtained from
the previous term: if the previous term is even, the next term is one half of the previous term. If the
previous term is odd, the next term is 3 times the previous term plus 1. The Collatz conjecture (also
called 3n + 1 conjecture) states that no matter what value of n, the sequence will always reach 1. For
example, if we start from 3, we get 10 as the second term, then 5, then 16, 8, 4, 2, 1, then we stop.
The sequence reaches 1 at the 8th term.
    (a) Write an R function Collatz() to calculate the number of steps to reach 1 for any positive integer.
    For example, you should get 1 for Collatz(1) and 8 for Collatz(3).
    (b) If $1 \le n \le 100$, which n maximizes Collatz(n)? How about $1 \le n \le 1000$?
    (c) Draw a scatter plot of the function Collatz(n) for $1 \le n \le 1000$.
    (d) Show the conjecture is not true if we start from a negative number, say `-17`. Construct a sequence
    from `-17` and show the sequence will return to `-17`. Print the whole cycle starting from `-17`.
    (e) How many different cycles can you find that involve a number -100 or above? The generalized
    Collatz conjecture is the assertion that the procedure, starting with any integer, eventually falls
    into one of the five cycles (1) 0 → 0; (2) 1 → 4 → 2 → 1; (3)`-1` → `-2` → `-1`; (4)`-5` → `-14` →
    `-7` → `-20` → `-10` → `-5`; (5)`-17` → `-50` → `-25` → `-74` → `-37` → `-110` → `-55` → `-164` →
    `-82` → `-41` → `-122` → `-61` → `-182` → `-91` → `-272` → `-136` → `-68` → `-34` → `-1`.

```{r}
collatz_helper <- function(x) {
  if ((x %% 2) == 0) {
    x <- x / 2
  } else if ((x %% 2) == 1) {
    x <- x * 3 + 1
  }
  return(x)
}



collatz <- function(num) {
  steps <- 0
  array <- numeric()
  orig <- num

  end_collatz <- c(0, 1, -1, -5, -17)

  # Edge cases 0 and 1
  if (orig == 0) {
    array <- c(array, num)
    return(list(steps = 0, intial = orig, values = unlist(array)))
  }

  if (orig == 1) {
    num <- collatz_helper(num)
    array <- c(array, num)
    steps <- steps + 1
    while (num != 1) {
      num <- collatz_helper(num)
      array <- c(array, num)
      steps <- steps + 1
    }
    return(list(steps = steps, intial = orig, values = unlist(array)))
  }

  # Regular case for positive integers
  if ((num > 0) && (num != 1)) {
    while (num != 1) {
      num <- collatz_helper(num)
      array <- c(array, num)
      steps <- steps + 1
    }
  }

  # Modified case for negative integers
  if ((num < 0) && (length(array) == 0)) {
    array <- c(array, num)
    num <- collatz_helper(num)
    array <- c(array, num)
    steps <- steps + 1
  }

  # Regular case for negative integers
  if ((num < 0) && (length(array) >= 1)) {
    while (!num %in% end_collatz) {
      num <- collatz_helper(num)
      array <- c(array, num)
      steps <- steps + 1
    }
  }
  return(list(steps = steps, intial = orig, values = unlist(array)))
}


max_collatz <- function(start, end) {
  step_size <- numeric()
  for (i in seq(start, end)) {
    col <- collatz(i)
    step_size <- c(step_size, col$steps)
  }
  return(list(n_val = c(start:end), num_steps = step_size))
}

# --------------------------------- Questions ---------------------------------
# (a)
print(collatz(8))

# (b)
max_100 <- max_collatz(1, 100)
cat(
  "The n value that maximizes collatz from 1 <= n <= 100 is",
  which.max(max_100$num_steps)
)
max_1000 <- max_collatz(1, 1000)
cat(
  "The n value that maximizes collatz from 1 <= n <= 1000 is",
  which.max(max_1000$num_steps)
)

# (c)
plot(max_1000$n_val, max_1000$num_steps,
  type = "l", xlab = "n",
  ylab = "collatz steps", lwd = 0.5, col = "#709d94",
  main = "Number of steps for Collatz() to converge"
)

# (d)
tmp_val <- paste(collatz(-17)$values, collapse = " → ")
cat(paste(strwrap(tmp_val, width = 70), collapse = "\n"))

# (e)
for (i in seq(-101, -110)) {
  col <- collatz(i)
  cat("------------------------ N:", i, "-----------------------\n")
  tmp_val <- paste(col$values, collapse = " → ")
  cat(paste(strwrap(tmp_val, width = 70), collapse = "\n"))
  cat("\n\n")
}
```

2. This question is about number theory. We will see how the problem becomes more challenging when
the scale goes up.
    (a) Create a function prime check that checks whether an input integer is prime or not.
    (b) Is 1234567891 a prime number? Is 1234567891234567891 a prime number? Is the answer from R
    reliable? (Try 1234567891234567891%%2) Does chatGPT give you a correct answer?
    (c) Create a function prime num that lists all prime numbers up to an input integer n. (It will be
    slow to use prime check. Check sieve of Eratosthenes in 375 notes chapter 4.)
    (d) Many even numbers can be written as a sum of two prime numbers. For example, 4 = 2 + 2,
    6 = 3 + 3, 8 = 3 + 5, 10 = 3 + 7 = 5 + 5, etc. Create a function prime sum that calculates
    how many ways an even integer can be written as a sum of two prime numbers. For example,
    prime sum(8) will return 1 and prime sum(10) will return 2. Find prime sum(400004).
    (e) Draw a scatter plot for the function in (c). (This is related to Goldbach`s conjecture.)
    (f) A twin prime is a pair of primes (x, y) with y = x + 2. Create a function prime twin that 
    calculates the number of twin prime pairs up to an integer n. Draw a scatter plot.

```{r}
prime_check <- function(n) {
  if (n <= 1) {
    return(FALSE)
  }
  if (n <= 3) {
    return(TRUE)
  }
  if (n %% 2 == 0 || n %% 3 == 0) {
    return(FALSE)
  }
  i <- 5
  while (i * i <= n) {
    if (n %% i == 0 || n %% (i + 2) == 0) {
      return(FALSE)
    }
    i <- i + 6
  }
  return(TRUE)
}

# (b) - part 1
b1 <- prime_check(1234567891)
if (b1 == TRUE) {
  print("1234567891 is a prime number")
} else {
  print("1234567891 is not a prime number")
}

# (b) - part 2
b2 <- prime_check(1234567891234567891)
if (b2 == TRUE) {
  print("1234567891234567891 is a prime number")
} else {
  print("1234567891234567891 is not a prime number")
}

# (c) - Sieve of Eratosthenes
prime_num <- function(n) {
  if (n > 2) {
    sieve <- seq(2, n)
    primes <- c()
    for (i in 2:n) {
      if (any(sieve == i)) {
        primes <- c(primes, i)
        sieve <- sieve[(sieve %% i) != 0]
      }
    }
    return(primes)
  } else {
    stop("Input value of n should be at least 2")
  }
}

print(prime_num(20))

# (d)
goldbach_conjecture <- function(even_num, pout = FALSE) {
  if (even_num <= 2 || even_num %% 2 != 0) {
    return("Number must be even and greater than 2.")
  }
  c <- NULL
  sol <- NULL
  for (i in 2:(even_num / 2)) {
    if (prime_check(i) && prime_check(even_num - i)) {
      c <- cbind(c, i) # nof solutions
      sol <- rbind(sol, data.frame(
        pair_1 = i,
        pair_2 = even_num - i
      ))
    }
  }
  if (pout) {
    cat("The number of Goldbach's pairs for", even_num, "is", length(c), "\n")
    cat("Some of the Goldbach's pairs for", even_num, "are\n")
    print(head(sol, 5))
  }
  return(length(c))
}

goldbach_conjecture(400004, TRUE)

# (e)
# calculate some 1000 solutions
sol <- NULL
for (i in seq(4, 1000, by = 2)) {
  nof_solutions <- goldbach_conjecture(i)
  sol <- rbind(sol, data.frame(n = i, nof = nof_solutions))
}
# plot solutions; alternating solutions
plot(sol$n, sol$nof,
  type = "p", xlab = "Even number", col = "#00d5ff",
  ylab = "Number of Solutions", main = "Goldbach's Conjecture"
)
reg <- lm(nof ~ n, data = sol)
abline(reg, col = "red")

# (f)
prime_twin <- function(n, pout = FALSE) {
  data <- 1:n
  prime_index <- sapply(data, prime_check)
  all_primes <- data[prime_index]
  twin2 <- all_primes[2:length(all_primes)][diff(all_primes == 2)]
  twin1 <- twin2 - 2
  bnd_cols <- cbind(twin1, twin2)
  if (pout) {
    cat("The number of twin pairs for", n, "is", length(bnd_cols), "\n")
    cat("Some of the twin pairs for", n, "are\n")
    df <- data.frame(bnd_cols)
    print(head(df, 5))
  }
  return(length(bnd_cols))
}

prime_twin(1000, TRUE)

sol <- NULL
for (i in seq(1, 1000)) {
  nof_solutions <- prime_twin(i)
  sol <- rbind(sol, data.frame(n = i, nof = nof_solutions))
}

# plot solutions; alternating solutions
plot(sol$n, sol$nof,
  type = "p", xlab = "Number", col = "#1e9155",
  ylab = "N_Pairs", main = "Number of twin primes"
)
```


3. Maximum subarray problem. Given X1, . . . , Xn, the maximum subarray problem solves
    $$
    S = \max_{1 \leq i < j \leq n} \sum_{k=i+1}^{j} X_k
    $$
    Implement several algorithms solving the maximum subarray problem. Compare the computation
    time.

```{r}
max_sum_subarr_nn <- function(arr) {
  # n^2 loop
  n <- length(arr)
  mx <- -Inf # Initialize maximum value

  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      sum <- sum(arr[i:j])
      mx <- max(mx, sum) # Update maximum value if needed
    }
  }

  if (mx > 0) {
    return(round(mx, 3))
  } else {
    return(0)
  }
}

max_sum_subarr_dp <- function(arr) {
  # Dynamic Programming
  current_sum <- arr[1]
  max_sum <- arr[1]

  # Loop from VALUE at index position 2 till the end of the array
  for (element in arr[-1]) {
    # If else statement executes faster than using max
    # current_sum <- max(current_sum + element, element)
    # max_sum <- max(current_sum, max_sum)

    if (current_sum + element < element) {
      current_sum <- element
    } else {
      current_sum <- current_sum + element
    }

    if (max_sum < current_sum) {
      max_sum <- current_sum
    } else {
      max_sum <- max_sum
    }
  }

  if (max_sum > 0) {
    return(max_sum)
  } else {
    return(0)
  }
}

# --------------------------------- Unit tests ---------------------------------
unittest_mx_sum_sub <- function(func) {
  # Case 1: Regular array
  input1 <- c(-1, 2, 3, -4, 2)
  expected_output1 <- 5
  test1 <- (func(input1) == expected_output1)

  # Test Case 2: All negative numbers
  input2 <- c(-3, -7, -2, -8, -6)
  expected_output2 <- 0
  test2 <- (func(input2) == expected_output2)

  # Test Case 3: All positive numbers
  input3 <- c(5, 7, 2, 8, 6)
  expected_output3 <- 28
  test3 <- (func(input3) == expected_output3)

  # Test Case 4: Regular array
  input4 <- c(-2, -3, 4, -1, -2, 1, 5, -3)
  expected_output1 <- 7
  test4 <- (func(input4) == expected_output1)

  if (test1 && test2 && test3 && test4) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

print(unittest_mx_sum_sub(max_sum_subarr_dp))
print(unittest_mx_sum_sub(max_sum_subarr_nn))


# ----------------------------- Benchmark tests -----------------------------
small_set <- rnorm(1e2)
med_set <- rnorm(1e3)
small_bench <- microbenchmark(
  max_sum_subarr_dp(small_set),
  max_sum_subarr_nn(small_set)
)
med_bench <- microbenchmark(
  max_sum_subarr_dp(med_set),
  max_sum_subarr_nn(med_set)
)
all_bench <- bind_rows(small_bench, med_bench)
print(all_bench)
```


4. Normalized maximum subarrary problem. Given X1, . . . , Xn, the maximum subarray problem solves
    $$T = \max_{1 \le i < j \le n} \sum_{k=i+1}^j \frac{X_k}{\sqrt{j-i}}$$
    Can you find a fast algorithm solving it?

```{r}
# O(n^2) complexity
max_norm_sum_subarr <- function(arr) {
  n <- length(arr)
  mx <- -Inf # Initialize maximum value

  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      subarr <- arr[i:j]
      sum <- sum(subarr) / sqrt(j + 1 - i)
      mx <- max(mx, sum) # Update maximum value if needed
      # cat(i, j, sqrt(j - i), sum, "\n")
    }
  }

  if (mx > 0) {
    return(round(mx, 3))
  } else {
    return(0)
  }
}

# --------------------------------- Unit tests ---------------------------------
unittest_mx_nm_sum_sub <- function(func) {
  # Case 1: Regular array
  input1 <- c(-1, 2, -3, 5, -4, 2)
  expected_output1 <- 2.309
  test1 <- (func(input1) == expected_output1)

  # Test Case 2: All negative numbers
  input2 <- c(-3, -7, -2, -8, -6)
  expected_output2 <- 0
  test2 <- (func(input2) == expected_output2)

  # Test Case 3: All positive numbers
  input3 <- c(5, 7, 2, 8, 6)
  expected_output3 <- 12.522
  test3 <- (func(input3) == expected_output3)

  if (test1 && test2 && test3) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

print(unittest_mx_nm_sum_sub(max_norm_sum_subarr))
```


5. Implement a version of `cumsum`. Can your function beat cumsum in computation time?

```{r}
cumsum1 <- function(xarr) {
  total <- rep(0, length(xarr))
  sum_tot <- 0
  for (i in seq(1, length(xarr))) {
    sum_tot <- sum_tot + xarr[i]
    total[i] <- sum_tot
  }
  return(total)
}


cumsum2 <- function(xarr) {
  total <- rep(0, length(xarr))
  total[1] <- xarr[1]
  for (i in seq(2, length(xarr))) {
    total[i] <- total[i - 1] + xarr[i]
  }
  return(total)
}


# Create a C++ implementation for cumulative sum
cppFunction("std::vector<double> cumsum_cpp(const std::vector<double>& data) {
  std::vector<double> result(data.size());

  // Initialize the first element of the result vector
  result[0] = data[0];

  // Calculate the cumulative sum using a simple for loop
  for (int i = 1; i < data.size(); ++i) {
    result[i] = result[i - 1] + data[i];
  }

  return result;
}")

# -------------------- A3P5 implementation --------------------
stat_675_cumsum <- function(x) {
  if (length(x) == 1) {
    return(x)
  }
  sx <- x
  for (i in 2:length(x)) {
    sx[i] <- sx[i] + sx[i - 1]
  }
  return(sx)
}

stat_675_cumsum2 <- function(x) {
  if (length(x) == 1) {
    return(x)
  }
  sx <- x
  x0 <- x[1]
  for (i in 2:length(x)) {
    x0 <- x0 + sx[i]
    sx[i] <- x0
  }
  return(sx)
}

stat_675_cumsum3 <- function(x) {
  n <- length(x)
  if (n < 10) {
    return(stat_675_cumsum2(x))
  }
  s <- floor(sqrt(n))
  X <- matrix(c(x, rep(0, s * (s + 2) - n)), s, s + 2)
  for (i in 2:s) {
    X[i, ] <- X[i, ] + X[i - 1, ]
  }
  for (i in 2:(s + 2)) {
    X[, i] <- X[, i] + X[s, i - 1]
  }
  return(as.vector(X)[1:n])
}

stat_675_cumsum4 <- function(x) {
  n <- length(x)
  if (n < 10) {
    return(stat_675_cumsum2(x))
  }
  s <- floor(sqrt(n))
  X <- matrix(c(x, rep(0, s * (s + 2) - n)), s + 2, s, byrow = TRUE)
  for (i in 2:s) {
    X[, i] <- X[, i] + X[, i - 1]
  }
  y <- stat_675_cumsum2(X[, s])
  X[-1, ] <- X[-1, ] + y[-(s + 2)]
  return(as.vector(t(X))[1:n])
}

# -------------------- Create arrays for tests --------------------
test_arr_sml <- c(-10:20)
test_arr_med <- c(-1e2:1e2)
test_arr_big <- c(-1e4:1e4)

# Units Tests
if (all.equal(cumsum(test_arr_sml), cumsum1(test_arr_sml), tolerance = 0)) {
  print("Passed Tests")
}

# Speed test
small_bench <- microbenchmark(
  cumsum(test_arr_sml), cumsum1(test_arr_sml),
  cumsum2(test_arr_sml), cumsum_cpp(test_arr_sml),
  stat_675_cumsum(test_arr_sml),stat_675_cumsum2(test_arr_sml),
  stat_675_cumsum3(test_arr_sml), stat_675_cumsum4(test_arr_sml)
)
med_bench <- microbenchmark(
  cumsum(test_arr_med), cumsum1(test_arr_med),
  cumsum2(test_arr_med), cumsum_cpp(test_arr_med),
  stat_675_cumsum(test_arr_med),stat_675_cumsum2(test_arr_med),
  stat_675_cumsum3(test_arr_med), stat_675_cumsum4(test_arr_med)
)
big_bench <- microbenchmark(
  cumsum(test_arr_big), cumsum1(test_arr_big),
  cumsum2(test_arr_big), cumsum_cpp(test_arr_big),
  stat_675_cumsum(test_arr_big),stat_675_cumsum2(test_arr_big),
  stat_675_cumsum3(test_arr_big), stat_675_cumsum4(test_arr_big)
)

all_bench <- bind_rows(small_bench, med_bench, big_bench)
print(all_bench)
```